A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-03-19 paper/paper.tex
Addressing the question of the contribution of the paper and of the stepwise explanation of the construction

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both augmented semi-simplicial and semi-cubical sets generated by iterating parametricity. In contrast to the standard presheaf presentation of augmented semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of augmented semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of augmented semi-simplices or semi-cubes in dimension $n$, we consider families of augmented semi-simplices and semi-cubes indexed over their faces.

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}
\subsection{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the connected components are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over a singleton set of colors.

@@

\section{Our construction}
In this section, we gradually build up an intuition for our construction, followed by the details.

@@

\subsection{Formal construction in ETT}
% Abbreviated tables in this section
%
% The rule for implicit arguments for each table:
% show 1 explicit in first two parts
% show E + 1 explicit in the third part
\appendmask[bonak]{layer}[D]
\appendmask[bonak]{painting}[D]
\appendmask[bonak]{restrframe}[D]
\appendmask[bonak]{restrlayer}[D, d]
\appendmask[bonak]{restrpainting}[D, d]
\appendmask[bonak]{cohframe}[D]
\appendmask[bonak]{cohlayer}[D, d]
\appendmask[bonak]{cohpainting}[D, d]

>>>>

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both augmented semi-simplicial and semi-cubical sets, generated by iterating the parametricity translation for type judgements. In contrast to the standard presheaf presentation of augmented semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of augmented semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of augmented semi-simplices or semi-cubes in dimension $n$, we consider families of augmented semi-simplices and semi-cubes indexed over their faces.

As a result, our work can be seen as contributing the following: it characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions; %Second, it makes fully formal the relation between semi-simplicial and semi-cubical sets and parametricity.
it provides with an alternative case study in the realm of indexed constructions of semi-simplicial and semi-cubical sets~\cite{voevodsky12,herbelin15,part15}, especially in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\cite{part15,shulman14,altenkirch16,kraus2017space}, as well as of better understanding the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett}); it makes a step in the direction of the program initiated in~\cite{altenkirch15} to develop parametricity-based models of parametric type theory~\cite{bernardy15,nuyts17,cavallo19} and cubical type theory~\cite{bezem13,cohen16,angiuli21} which are closer from the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do (compare e.g. to the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally closed cartesian categories~\cite{curien14}).

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}

We give in this section a uniform definition of semi-simplicial and semi-cubical sets which we believe is folklore. The uniformity allows the definition of a notion of $\nu$-sets that subsumes semi-simplicial and semi-cubical sets.

\subsection{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the connected components are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over a singleton set of colors.

@@

\section{Our construction}

Our construction involves several details, some of them being visible only in the Coq formalization. A description at an informal level which we estimate to be however mathematically precise enough can be found in Sections~\ref{sec:ett} and~\ref{sec:itt}, while technical issues with the mechanization are explained in~\ref{sec:mech}.

The whole definition is highly dependent, with dependencies in
inequality proofs and in coherence conditions. In Section~\ref{sec:ett} and Tables~\ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces}, \ref{tab:coh}, we
give a definition in Extensional Type Theory, which roughly
corresponds to a level of reasoning ``à la set theory'' where coherence
conditions are thought as ``holding on the nose'' without having to
mention them explicitly once they are justified. In Section~\ref{sec:itt} and Tables~\ref{fulltab:coind}, \ref{fulltab:core}, \ref{fulltab:frames}, \ref{fulltab:faces}, \ref{fulltab:coh}, we
give a definition in Intentional Type Theory, where transporting along
the coherence conditions is made explicit. This makes the definition
more involved, but at the same time helps to understand the
specificities of coherence conditions.

But first, we start with a section gradually and informally explaining the intuition of the construction.

@@

\subsection{Formal construction in ETT\label{sec:ett}}
% Abbreviated tables in this section
%
% The rule for implicit arguments for each table:
% show 1 explicit in first two parts
% show E + 1 explicit in the third part
\appendmask[bonak]{layer}[D]
\appendmask[bonak]{painting}[D]
\appendmask[bonak]{restrframe}[D]
\appendmask[bonak]{restrlayer}[D, d]
\appendmask[bonak]{restrpainting}[D, d]
\appendmask[bonak]{cohframe}[D]
\appendmask[bonak]{cohlayer}[D, d]
\appendmask[bonak]{cohpainting}[D, d]
